\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs

### 2023-10-14 paper.tex
Updates from Yan and gregor

% What do we do?
In this paper, we attempt to address these problems by
providing a precise definition of an ARG that is general enough to encompass
the output of popular simulation and inference methods.
We propose a definition of genetic genealogy in terms of specific genomes
and their intervals of genetic inheritance, and refer to this as
a ``genome ARG'', or ``gARG''.
This contrasts with the classical Griffiths approach,
based on recording evolutionary events, which we call
an ``event ARG'', or ``eARG''.
We show that the gARG encoding has a range of advantages over the
eARG approach, being flexible, efficient and, in particular,
allowing us to systematically encode uncertainty about the
the precise details of recombination events.

TODO: Provide a map of the paper's sections and how they tie together in support of gARG as a way forward.

@@

A genome ARG (gARG) is a directed acyclic graph in which nodes represent
haploid genomes and edges represent
genetic inheritance between an ancestor and a descendant.
The topology of a gARG specifies that genetic inheritance
occurred between particular
ancestors and descendants, but the graph connectivity
does not tell us which \emph{parts} of their genomes were inherited.
In order to capture the effects of recombination
(and other processes that lead to the partial genome inheritance)
 we ``annotate'' the edges with the genome
coordinates over which inheritance occurs.
We can define a gARG formally as follows.
Let $N \subset \mathbb{N}$ be the set of nodes representing
the genomes in the gARG, and $E$ be the set of edges.
We will usually define a set of nodes $S \subseteq N$
representing the sampled genomes
(but see section XXX for some subtlety around this point).
Each element of $E$ is a tuple $(c, p, I)$ such that $c, p \in N$ are the child and
parent nodes and $I$ is the set of disjoint genomic intervals
over which genome $c$ inherits from $p$.
Thus, each topological connection between
a parent and child node in the graph is ``annotated'' with a set of
inheritance intervals $I$.
Here, the terms parent and child are used in graph sense;
these nodes can respectively represent an ancestor and a descendant genome,
which can be genetically separated by multiple meioses.
We will use these two sets of terms interchangeably.

In many settings, nodes are dated, i.e.\
each node $u\in N$ is associated with a time $\tau_u$.
However, the topological relationships
defined by the directed graph structure is sufficient for
many applications, and we will assume that node dates are not
known here.
% This should be obvious, but seems like it needs to be said
In practical settings, we will wish to associate additional
metadata such as a sample identifiers or quality-control metrics
with nodes; it is therefore best to think of the
integers used here in the definition of a node as an \emph{identifier},
with which arbitrary additional information can be associated.

@@

As illustrated in Fig.~\ref{fig-arg-in-pedigree},
the gARG for a given set of individuals is embedded in their pedigree.
The pedigree relationship between the diploid individuals is
shown, along with their paths
of genetic inheritance and constituent genomes (labelled
using using lowercase alphabetical letters
rather than integer identifiers to avoid confusion with genomic
intervals).
Thus, individual $D_1$ is composed
of genomes \noderef{a} and \noderef{b}, which are inherited from its
two parents, $D_3$ and $D_4$. Each inherited genome may be the recombined product
of the genomes contained within a parent. In this example,
genome \noderef{b} was inherited directly from $D_4$'s genome \noderef{g} without
recombination, whereas
genome \noderef{a} is the recombinant of
$D_2$'s genomes \noderef{e} and \noderef{f} crossing over at position 2.
Specifically, genome \noderef{a} inherited the (half-closed)
interval $[0, 2)$ from genome \noderef{e} and $[2, 10)$ from genome \noderef{f}.
These intervals are shown attached to the corresponding graph edges.
The figure shows the annotated pedigree with realised inheritance of genomes
between generations, the corresponding gARG, and finally the corresponding
sequence of local trees along the genome.
The local trees represent the genealogy of genomes \noderef{a}--\noderef{d},
which changes between local trees due to recombination.
The local trees span the three genome regions delineated
by the two recombinations that gave rise to these genomes.
% Following the above definitions, corresponding data for this gARG is shown in Supplementary? Table XXX.

@@

To make the following discussions precise, we require
a concrete definition and explicit process for local tree extraction.
We can describe a local tree as an
oriented tree~\cite[p.\ 461]{knuth11combinatorial},
a sequence of integers $\pi_1, \pi_2\dots$ where $\pi_u$ denotes the parent of
node $u$~\citep{kelleher2013coalescent,kelleher2014coalescent,kelleher2016efficient}.
If $u$ is a root
% I'm not sure if this reachability stuff is a helpful clarification, and whether it'll just
% confuse people. It depends on some tricky details.
(or not reachable from the samples)
in the local tree at position $x$, then $\pi_u = -1$ (by convention).
% This sentence needs another pass
A node in an oriented tree can have
zero, one (see the next section for discussion of ``unary'' nodes),
or many children.
To recover the local tree $\pi$ at position $x$, we begin by
setting $\pi_u = -1$ for each $u\in N$. Then, for each sample
node in $S$ we trace its path rootwards through the
ARG for position $x$, and record this path in $\pi$.
Specifically, at a given node $u$,
we find an edge $(c, p, I) \in E$ such that $u = c$ and $x \in I$, and set
$\pi_c \leftarrow p$. We then set $u \leftarrow p$, and repeat
until either $\pi_u \neq -1$ (indicating we have traversed this section
of the ARG already on the path from another sample) or there
is no matching outbound edge (indicating we are at a root).

@@

This is not the prevailing view, however.
\cite{kuhner2017consensus} argue that the
``interval-tree'' representation of
an ARG (the local trees and the genome intervals they cover)
``does not contain all of the information in the underlying ARG: it lacks the
number of recombinations occuring at each site, the times at which
recombinations occurred, and the specific sequences involved as recombination
partners.''
\cite{shipilina2023origin} discuss the same ideas,
and note that the
``full ARG...~contains more information than the series of tree
sequences along the genome''.
These statements that an ARG
contains more information than its local trees is only true
if we use a representation of the local trees that discards
information about those local trees that can be derived from the ARG.
The internal nodes of evolutionary trees are usually
considered to be \emph{unlabelled}
because they most often correspond
to anonymous hypothetical ancestors.
Internal ARG nodes may likewise correspond to anonymous hypothetical
ancestors, but if we do not label the tree nodes we lose the fact that
the \emph{same} ancestors persist across multiple trees.
Besides having labelled nodes, the local trees must contain
unary nodes (see the next section) marking recombinations
and other processes that lead to lineages passing through a given
ARG node without coalescence in the local trees (see section XXX).
If the local trees contain labelled internal nodes, precisely
describing the corresponding path through the graph for each
genome interval, then there is exactly as
much information in them as in an ARG.

@@

This distinction between graph and local tree arity is mainly
of interest when we consider \emph{unary} nodes, those that have
a single child.
Returning to the example in Fig~\ref{fig-arg-in-pedigree}, nodes
\noderef{g} and \noderef{h} are ARG-unary (Fig~\ref{fig-arg-in-pedigree}B), and are consequently
also unary in the local trees (Fig~\ref{fig-arg-in-pedigree}C).
On the other hand, node \noderef{f} has two children
in the graph, but is binary only
in the local tree covering the interval $[2, 7)$,
representing the coalescence of samples \noderef{a} and \noderef{c}
in this region of genome. Over the interval $[0, 2)$ no coalescence occurs,
but we still record the fact that genome \noderef{c} inherits from \noderef{f}
in the local tree. Thus, \noderef{c} has a single child in this
interval and it is hence unary in this local tree.
In another example, \noderef{e} is binary in the graph, but is
unary in each of the local trees it is present in.
This is because \noderef{a} and \noderef{c} are both recombinant
offspring of \noderef{e} and \noderef{f}, and,
in this example,
\noderef{a} inherits from
\noderef{e} over the interval $[0, 2)$,
and \noderef{c} inherits from \noderef{e} over $[7, 10)$.
The intervals over which
we know that a node inherits from another, but where there
is no local coalescence, are marked by unary nodes in the
local trees.

@@

\section*{Levels of simplification}\label{ARG_simplification}
ARG simplification is a powerful tool, and can do far more
than removing unreachable nodes and edges in a prospective ARG
and resolving ancestral material in an eARG.
In general, we can think of
ARG simplification as the process
of removing nodes and re-writing edges (and their inheritance annotations)
to remove various types of redundancy.
The redundancy that we are interested in
% We are interested in both graph and local unary, so qualifying this
% would just make it less readable, IMO
revolves around the presence unary nodes (see previous section).
We illustrate this successive removal of redundancy
through a series simplification steps
in Fig.~\ref{fig-simplification}.

@@

Fully decoupling the general concept of an ARG from the coalescent
with recombination stochastic process is an important step.
While this has proven to be a useful and
robust
model~\citep{wakeley2012gene,bhaskar2014distortion,nelson2020accounting},
many modern datasets have properties that grossly
violate its assumptions.
For instance, several human datasets now consist of hundreds of thousands of
% TODO cite GeL paper
genomes~\citep{bycroft2018genome,karczewski2020mutational,tanjo2021practical},
and so sample size is substantially \emph{larger} than the
usually assumed $N_e$ values.
Agricultural datasets are an even more extreme departure from coalescent
assumptions, with hundreds of thousands of samples embedded in
multi-generational pedigrees~\citep{hayes20191000,Ros-Freixedes2020}
and effective population sizes of 100 and even
less~\citep{MacLeod2013,Makanjuola2020,Hall2016,Porcnic2016}.
Even if sufficiently complex demographic models~\citep{gower2022demes}
encompassing hundreds of populations, explosive growth rates and
myriad interconnections of migration, were somehow estimated and
provided as input, ARGs sampled from
the coalescent cannot capture the complexities of realistic family structures. There are many species with sizeable to large number of progeny, making polytomies a necessary feature of local trees. Flexible ARG encoding should accommodate such cases.

@@

% JK: I'd usually use "JK was supported by XX", but this form seems fine too?
% I don't have an opinion about it, if others think something else is more
% appropriate.
GG acknowledges support from the BBSRC ISP grant to The Roslin Institute (BBS/E/D/30002275, BBS/E/RL/230001A, BBS/E/RL/230001C).
Jerome Kelleher acknowledges support from the Robertson Foundation.
Jere Koskela was supported by EPSRC research grant EP/V049208/1.

@@

At the base of this concept is a subtle and important point about which genomes the nodes in a gARG represent.
In principle, we can interpret nodes as representing genomes
at any point in the life cycle, for instance, \citet{hudson1983properties} views nodes as representing gametes,
whereas in Fig.~\ref{fig-arg-in-pedigree}A, we present them as genomes within a diploid individual.
In reality, however, recombination and replication of DNA occur at a cellular level,
and correspond to the processes of meiosis and mitosis respectively. A comprehensive gARG would therefore
trace the process of inheritance not between diploid individuals, but between each cell during the
life cycle of an organism. This is illustrated in Fig.~\ref{fig-cell-lines}, where we show a simplified
schematic of cellular inheritance in a diploid population over four partially overlapping generations.
In this figure, thick black lines trace an ARG that describes the ancestry of the lower arm of the
four sampled chromosomes. The ``events'' underlying the ARG are shown as red and blue squares,
and the potential ARG nodes are shown as green chromosomal regions. Note that a number of these nodes are
``pass-through'' nodes with one parent and one child, and in practice these would normally be omitted.

>>>>

% What do we do?
In this paper, we attempt to address these problems by
providing a precise definition of an ARG that is general enough to encompass
the output of popular simulation and inference methods.
We propose a definition of genetic genealogy in terms of specific genomes
and their intervals of genetic inheritance, and refer to this as
a ``genome ARG'', or ``gARG''.
This contrasts with the classical Griffiths approach,
based on recording evolutionary events, which we call
an ``event ARG'', or ``eARG''.
We show that the gARG encoding has a range of advantages over the
eARG approach, being flexible, efficient and, in particular,
allowing us to systematically encode uncertainty about
the precise details of recombination events.

% TODO: Provide a map of the paper's sections and how they tie together in support of gARG as a way forward.
% Here is a rough draft - it would be good to state the most important message (in one sentence or so) from each of the sections to help with frontloading key messages
% Genome ARGs &
% Event ARGs
In the following, we define the genome ARG (gARG) and contrast it with the event ARGs (eARG).
% Retrospective vs prospective
% TODO: This needs revision - see also https://github.com/tskit-dev/what-is-an-arg-paper/pull/380
We continue with a description of the usually held retrospective view of ARGs, but note that ARGs can also be viewed as prospective, particularly in forwards-time simulations. In the context of forwards-time simulations, we mention the simplification operation, which removes ancestral genome material that was not inherited.
% Converting and eARG to gARG &
% ARGs and local trees &
% Locally unary nodes &
% Levels of simplification
This simplification is closely related to operations needed to convert eARG and gARG, which opens a rich set of details about ARGs and local trees, local unary nodes, and levels of simplification.
% Precision of recombination information
We then highlight the important difference between the required level of precision about recombination information for eARGs compared to gARGs.
% Computational efficiency and data interchange
The way this recombination information is stored in eARGs and gARGs has important implications for computational efficiency, which is critical if we want to use ARGs for modern genomic datasets.
% Discussion
We wrap with a discussion highlighting that WHAT-IS-THE-KEY-SINGLE-MESSAGE-NOT-COVERED-ABOVE? Finally, the literature on ARGs and associated aspects is vast, particularly due to different definitions and implementations. We cover some important aspects in appendices, including a brief history of ancestral graphs (Appendix XXX), description of the big and little ARGs (Appendix ZZZ), survey of ARG inference methods (Appendix YYY), and ARGs at an individual vs cell lineage level (Appendix WWW).

@@

A genome ARG (gARG) is a directed acyclic graph in which nodes represent
haploid genomes and edges represent
genetic inheritance between an ancestor and a descendant.
The topology of a gARG specifies that genetic inheritance
occurred between particular
ancestors and descendants, but the graph connectivity
does not tell us which \emph{parts} of their genomes were inherited.
In order to capture the effects of recombination
(and other processes that lead to the partial genome inheritance)
 we ``annotate'' the edges with the genome
coordinates over which inheritance occurs.
We can define a gARG formally as follows.
Let $N \subset \mathbb{N}$ be the set of nodes representing
the genomes in the gARG, and $E$ be the set of edges.
We will usually define a set of nodes $S \subseteq N$
representing the sampled genomes
(but see section XXX for some subtlety around this point).
Each element of $E$ is a tuple $(c, p, I)$ such that $c, p \in N$ are the child and
parent nodes and $I$ is the set of disjoint genomic intervals
over which genome $c$ inherits from $p$.
Thus, each topological connection between
a parent and child node in the graph is ``annotated'' with a set of
inheritance intervals $I$.
Here, the terms parent and child are used in graph sense;
these nodes can respectively represent ancestor and descendant genomes,
which can be separated by multiple generations.
We will use these two sets of terms interchangeably.

In many settings, nodes are dated, i.e.\
each node $u\in N$ is associated with a time $\tau_u$.
However, the topological relationships
defined by the directed graph structure is sufficient for
many applications, and we will assume that node dates are not
known here.
% This should be obvious, but seems like it needs to be said
In practical settings, we will wish to associate additional
metadata such as sample identifiers or quality-control metrics
with nodes; it is therefore best to think of the
integers used here in the definition of a node as an \emph{identifier},
with which arbitrary additional information can be associated.

@@

As illustrated in Fig.~\ref{fig-arg-in-pedigree},
the gARG for a given set of individuals is embedded in their pedigree.
The pedigree relationship between the diploid individuals is
shown, along with their paths
of genetic inheritance and constituent genomes (labelled
with lowercase alphabetical letters
rather than integer identifiers to avoid confusion with genomic
intervals).
Thus, individual $D_1$ is composed
of genomes \noderef{a} and \noderef{b}, which are inherited from its
two parents, $D_3$ and $D_4$. Each inherited genome may be the recombined product
of the genomes contained within a parent. In this example,
genome \noderef{b} was inherited directly from $D_4$'s genome \noderef{g} without
recombination, whereas
genome \noderef{a} is the recombinant of
$D_2$'s genomes \noderef{e} and \noderef{f} crossing over at position 2.
Specifically, genome \noderef{a} inherited the (half-closed)
interval $[0, 2)$ from genome \noderef{e} and $[2, 10)$ from genome \noderef{f}.
These intervals are shown attached to the corresponding graph edges.
The figure shows the annotated pedigree with realised inheritance of genomes
between generations, the corresponding gARG, and finally the corresponding
sequence of local trees along the genome.
The local trees represent the genealogy of genomes \noderef{a}--\noderef{d},
which changes between local trees due to recombination.
The local trees span the three genome regions delineated
by the two recombination breakpoints that gave rise to these genomes.
% Following the above definitions, corresponding data for this gARG is shown in Supplementary? Table XXX.

@@

To make the following discussions precise, we provide
a concrete definition and explicit process for local tree extraction.
We can describe a local tree as an
oriented tree~\cite[p.\ 461]{knuth11combinatorial},
a sequence of integers $\pi_1, \pi_2\dots$ where $\pi_u$ denotes the parent of
node $u$~\citep{kelleher2013coalescent,kelleher2014coalescent,kelleher2016efficient}.
If $u$ is a root
% I'm not sure if this reachability stuff is a helpful clarification, and whether it'll just
% confuse people. It depends on some tricky details.
(or not reachable from the samples)
in the local tree at position $x$, then we adopt the convention that $\pi_u = -1$.
% This sentence needs another pass
A node in an oriented tree can have
zero, one,
or many children.
To recover the local tree $\pi$ at position $x$, we begin by
setting $\pi_u = -1$ for each $u\in N$. Then, for each sample
node in $S$ we trace its path rootwards through the
ARG for position $x$, and record this path in $\pi$.
Specifically, at a given node $u$,
we find an edge $(c, p, I) \in E$ such that $u = c$ and $x \in I$, and set
$\pi_c \leftarrow p$. We then set $u \leftarrow p$, and repeat
until either $\pi_u \neq -1$ (indicating we have traversed this section
of the ARG already on the path from another sample) or there
is no matching outbound edge (indicating we are at a root).

@@

This is not the prevailing view, however.
\cite{kuhner2017consensus} argue that the
``interval-tree'' representation of
an ARG (the local trees and the genome intervals they cover)
``does not contain all of the information in the underlying ARG: it lacks the
number of recombinations occuring at each site, the times at which
recombinations occurred, and the specific sequences involved as recombination
partners.''
\cite{shipilina2023origin} discuss the same ideas,
and note that the
``full ARG...~contains more information than the series of tree
sequences along the genome''.
These statements that an ARG
contains more information than its local trees is only true
if we use a representation of the local trees that discards
information about those local trees that can be derived from the ARG.
The internal nodes of evolutionary trees are usually
considered to be \emph{unlabelled}
because they most often correspond
to anonymous hypothetical ancestors.
Internal ARG nodes may likewise correspond to anonymous hypothetical
ancestors, but if we do not label the tree nodes we lose the fact that
the \emph{same} ancestors persist across multiple trees.
Besides having labelled nodes, the local trees must contain
unary nodes (see the next section) marking recombinations
and other processes that lead to lineages passing through a given
ARG node without coalescence in the local trees (see section XXX).
If the local trees contain labelled internal nodes, they precisely
describing the corresponding path through the graph for each
genome interval and there is exactly as
much information in them as in an ARG.

@@

This distinction between graph and local tree arity is mainly
of interest when we consider \emph{unary} nodes, those that have
a single child.
Returning to the example in Fig~\ref{fig-arg-in-pedigree}, nodes
\noderef{g} and \noderef{h} are ARG-unary (Fig~\ref{fig-arg-in-pedigree}B), and are consequently
also unary in the local trees (Fig~\ref{fig-arg-in-pedigree}C).
On the other hand, node \noderef{f} has two children
in the graph, but is binary only
in the local tree covering the interval $[2, 7)$,
representing the coalescence of samples \noderef{a} and \noderef{c}
in this region of genome. Over the interval $[0, 2)$ no coalescence occurs,
but we still record the fact that genome \noderef{c} inherits from \noderef{f}
in the local tree. Thus, \noderef{c} has a single child in this
interval and it is hence unary in this local tree.
In another example, \noderef{e} is binary in the graph, but is
unary in each of the local trees in which it is present.
This is because \noderef{a} and \noderef{c} are both recombinant
offspring of \noderef{e} and \noderef{f}, and,
in this example,
\noderef{a} inherits from
\noderef{e} over the interval $[0, 2)$,
and \noderef{c} inherits from \noderef{e} over $[7, 10)$.
The intervals over which
we know that a node inherits from another, but where there
is no local coalescence, are marked by unary nodes in the
local trees.

@@

\section*{Levels of simplification}\label{ARG_simplification}
ARG simplification is a powerful tool, and can do far more
than removing unreachable nodes and edges in a prospective ARG
and resolving ancestral material in an eARG.
In general, we can think of
ARG simplification as the process
of removing nodes and re-writing edges (and their inheritance annotations)
to remove various types of redundancy.
The redundancy that we are interested in
% We are interested in both graph and local unary, so qualifying this
% would just make it less readable, IMO
revolves around the presence of unary nodes (see previous section).
We illustrate this successive removal of redundancy
through a series simplification steps
in Fig.~\ref{fig-simplification}.

@@

Fully decoupling the general concept of an ARG from the coalescent
with recombination stochastic process is an important step.
While this has proven to be a useful and
robust
model~\citep{wakeley2012gene,bhaskar2014distortion,nelson2020accounting},
many modern datasets have properties that grossly
violate its assumptions.
For instance, several human datasets now consist of hundreds of thousands of
% TODO cite GeL paper
genomes~\citep{bycroft2018genome,karczewski2020mutational,tanjo2021practical},
and so sample size is substantially \emph{larger} than the
usually assumed $N_e$ values.
Agricultural datasets are an even more extreme departure from coalescent
assumptions, with hundreds of thousands of samples embedded in
multi-generational pedigrees~\citep{hayes20191000,Ros-Freixedes2020}
and effective population sizes of 100 and even
less~\citep{MacLeod2013,Makanjuola2020,Hall2016,Porcnic2016}.
Even if sufficiently complex demographic models~\citep{gower2022demes}
encompassing hundreds of populations, explosive growth rates and
myriad interconnections of migration, were somehow estimated and
provided as input, ARGs sampled from
the coalescent cannot capture the complexities of realistic family structures, which often involve overlapping generations and extreme variation in lifetime reproductive success. Flexible ARG encoding should accommodate such cases.

@@

% FIXME: simplest to order this alphabetically by surname?
% Anastasia TODO
Jere Koskela was supported by EPSRC research grant EP/V049208/1.
Gregor Gorjanc acknowledges support from the BBSRC ISP grant to The Roslin Institute (BBS/E/D/30002275, BBS/E/RL/230001A, and BBS/E/RL/230001C).
% Wilder TODO
Jerome Kelleher acknowledges support from the Robertson Foundation.

@@

At the base of this concept is a subtle and important point about which genomes are represented by gARG nodes.
In principle, we can interpret nodes as representing genomes
at any point in the life cycle, for instance, \citet{hudson1983properties} views nodes as representing gametes,
whereas in Fig.~\ref{fig-arg-in-pedigree}A, we present them as genomes within a diploid individual.
In reality, however, recombination and replication of DNA occur at a cellular level,
and respectively correspond to the processes of meiosis and mitosis. Therefore, a comprehensive gARG would not
trace genetic inheritance between diploid individuals, but between cells during the
life cycle of these individuals. This is illustrated in Fig.~\ref{fig-cell-lines}, where we show a simplified
schematic of cellular inheritance in a diploid population over four partially overlapping generations.
In this figure, thick black lines trace an ARG that describes the ancestry of the lower arm of the
four sampled chromosomes. The ``events'' underlying the ARG are shown as red and blue squares,
and the potential ARG nodes are shown as green chromosomal regions. Note that a number of these nodes are
``pass-through'' nodes with one parent and one child, which would normally be omitted.
